\vspace{1.0cm} \LARGE \bf Revealing the competitive dynamics of influenza viruses through evolutionary cartography

### 2012-12-17 flux.tex
Revise manuscript.

Before 2009, there were four major clades of influenza circulating within the human population: influenza A H3N2 and H1N1, and influenza B Victoria and Yamagata. 
In the case of influenza A, subtypes A/H3N2 and A/H1N1 refer to the genes, hemagglutinin (H or HA) and neuraminidase (N or NA), that are primarily responsible for the antigenic character of a strain. 
In the case of influenza B, B/Vic and B/Yam refer to antigenically distinct lineages which diverged from one prior to 1980 \cite{Rota92}.
Mutations to the HA1 region of the hemagglutinin protein are thought to drive the majority of antigenic drift in the influenza virus \cite{Nelson07NatRevGenet}. 
Experimental characterization of antigenic phenotype is possible through the hemagglutination inhibition (HI) assay \cite{Hirst43}, which measures the cross-reactivity of one virus strain to serum raised against another strain through challenge or vaccination. 
Sera from older strains react poorly with more evolved viruses resulting in new strains having a transmission advantage over established strains.
HI assays have shown there to be little to no cross-reactivity between influenza clades A/H3N2, A/H1N1, B/Vic and B/Yam \cite{Hay01}.

The results of many HI assays across a multitude of virus strains can be combined to yield a two-dimensional map, quantifying antigenic similarity and distance \cite{Smith04}. 
The antigenic map of influenza A (H3N2) has shown largely linear movement of the influenza virus population since its introduction in 1968. 
Evolution of antigenic phenotype appears punctuated with periods of stasis interspersed by periods of more rapid innovation, while genetic evolution appears more continuous \cite{Smith04}, suggesting that a relatively small number of genetic changes or combinations of genetic changes may drive changes in antigenic phenotype. 
The process of antigenic drift results in the rapid turnover of the virus population. 
Although mutation occurs rapidly, standing genetic diversity is low and phylogenetic analysis shows a characteristically `spindly' tree with a single predominant trunk lineage and transitory side branches that persist for only 1--5 years \cite{Fitch97}.

@@

In order to assess patterns of antigenic evolution among influenza strains, we implemented a Bayesian probabilistic analog of multidimensional scaling, referred to here as BMDS (see Methods).
In this model, viruses and sera are given $N$-dimensional locations, thus specifying an `antigenic map', such that distances between viruses and sera in this space are inversely proportional to cross-reactivity.
In the BMDS model, a map distance of one antigenic unit translates to an expectation of a 2-fold drop in HI titer between virus and sera.
Maps that produce pairwise distances most congruent with the observed titers will have a high likelihood and will be favored by the BMDS model.
We integrate over sources of uncertainty, such as antigenic locations, in a flexible Bayesian fashion.
We apply this model to HI measurements of virus isolates against post-infection ferret sera for influenza A/H3N2, A/H1N1, B/Vic and B/Yam (see Methods).

We test model performance by constructing a training dataset representing 90\% of the HI measurements of the A/H3N2 dataset and a test dataset representing the remaining 10\% of the full dataset. 
By fitting the BMDS model to the training dataset, we are able to predict HI titers in the test dataset and compare predicted titers to observed titers in the test dataset. 
We find that a BMDS analog of the model used by Smith et al.\ \cite{Smith04}, in which viruses and sera are represented as 2D locations and expected titer is relative to the maximum titer of a particular ferret sera, performs well, yielding an average absolute predictive error of 0.87 log$_2$ HI titers (Table \ref{errortable}).
We find that a model with two dimensional antigenic locations predicts test data to the same precision as a 3D model, but performs substantially better than a 1D model or models with four or more dimensions (Table \ref{errortable}), and thus specify a two dimensional model in all subsequent analyses.
We extend this model by estimating the strength of overall reactivity of each serum rather than fixing this at the maximum titer, and additionally, by estimating the strength of reactivity of each virus isolate.
We refer to these estimates as serum effects and virus effects, respectively.
We found that including these effects decreased test error to 0.75 for influenza A/H3N2.
We arrive at similar precisions when repeating this analysis for A/H1N1, B/Vic and B/Yam.
For the 2D model with serum and virus effects, A/H1N1 shows a test error of 0.64, B/Vic shows a test error of 0.74 and B/Yam shows a test error of 0.80.

@@

Most previous work on virus antigenic and genetic evolution has examined antigenic and genetic patterns in isolation, and then compared the results to assess congruence of these two aspects of evolution \cite{Hay01,Smith04,Russell08}. 
Here, we simultaneously model antigenic and genetic evolution by adapting an evolutionary diffusion process \cite{Lemey10}, wherein a virus's antigenic character state evolves along branches of phylogenetic tree according to Brownian motion process (see Methods).
Here, the diffusion process acts as a prior, so that genetically similar viruses are expected to share similar antigenic phenotypes, and we model the relationship between genetic change and accrual of antigenic distance.
We include sequence data for A/H3N2, A/H1N1, B/Vic and B/Yam to estimate this diffusion process.
We find similar levels of predictive accuracy when including the diffusion process with A/H3N2 showing a test error of 0.75, A/H1N1 showing a test error of 0.66, B/Vic showing a test error of 0.70 and B/Yam showing a test error of 0.78.

@@

Through our modeling, we reveal the evolutionary basis for antigenic phenotype in influenza A/H3N2, A/H1N1, B/Vic and B/Yam (Figure~\ref{map}).
Over the time period of 1968 to 2011, influenza A/H3N2 shows substantially more antigenic evolution than is exhibited by A/H1N1 over the course of 1977 to 2009 or B/Vic and B/Yam over the course of 1986 to 2011.
We observe prominent antigenic clusters in A/H3N2 and A/H1N1 with less obvious clustering in B/Vic and B/Yam.
Antigenic clusters show high genetic similarity, so that we only observe a single diffusion event leading to each cluster, rather than the repeated emergence of clusters.
This analysis makes the fate of antigenic clusters obvious, with two clusters in A/H3N2 (Texas/77 and Beijing/89) appearing to be evolutionary dead-ends.

@@

HI assays lack sensitivity beyond a certain point, so that for A/H3N2, cross-reactive measurements only exist between strains sampled at most 11 years apart, leaving only threshold titers, e.g.\ `$<$40', in more temporally distant comparisons.  
Because of the threshold of sensitivity of the HI assay, it's impossible to distinguish a linear trajectory in 2D antigenic space, from a curved trajectory, so long as the curve does not bring antigenic phenotype full circle to have cross-reactive measurements between temporally distant strains.
To solve this problem of identifiability, we assumed a weak prior that favors linear movement in the 2D antigenic space, with the slope of the linear relationship and the precision of the relationship incorporated into the Bayesian model (see Methods).
Because of this, we interpret antigenic locations locally rather than globally.
We can determine the rate of antigenic movement of virus lineages without knowing the larger configuration that the movement occurs under.

@@

We find that other clades of influenza evolved in antigenic phenotype substantially slower than A/H3N2 (Figure~\ref{drift}).
Influenza A/H1N1 evolved at a rate of 0.55 antigenic units per year (HPD 0.49--0.62), but showing a similar pattern of punctuated antigenic evolution with occasional larger jumps in phenotype, such as the emergence of the Solomon Islands/06 cluster.  
Influenza B/Victoria also evolved relatively slowly, with an average rates of 0.41 (HPD 0.29--0.55) antigenic units per year.
Influenza B/Vic appears to show some degree of punctuated antigenic evolution with a recent transition creating the Brisbane/08 cluster.
Interestingly, a minor lineage of B/Vic has persisted that rather than moving forward in antigenic dimension 1 has moved down in antigenic dimension 2 (Figure~\ref{map}, Figure~\ref{drift}).
Influenza B/Yamagata evolved slower still, with an average rate of and 0.21 (HPD 0.10--0.28) antigenic units per year.
Very little punctuated evolution is apparent in the evolution of B/Yam.

@@

Additionally, year-to-year comparisons show a similar pattern (Figure~\ref{incidence}). 
Here, we define antigenic drift from year $i-1$ to year $i$ as the difference between the average location in dimension 1 in viruses sampled in year $i-1$ and the average location in dimension 1 of viruses sampled in year $i$.
For instance, we see that 97\% of the antigenic drift between 1998 and 1999 is attributable to A/H3N2 and subsequently the 1999/2000 winter influenza season is dominated by A/H3N2.
Similarly, between samples from 1999 and 2000, most antigenic drift is attributable to A/H1N1 and B/Yam, and the 2000/2001 season is dominated by these same two clades.
Here, we find an overall correlation between clade-specific relative antigenic drift and clade-specific incidence of $r = 0.67$ (Pearson correlation, $p = 1.00 \times 10^{-6}$).
Correlation is present to greater of lesser degrees for each of the four clades, with $r_\mathrm{H3} = 0.63$, $r_\mathrm{H1} = 0.29$, $r_\mathrm{Vic} = 0.75$ and $r_\mathrm{Yam} = 0.58$.

@@

Here, we examine the conversion of newly acquired antigenic mutation into fixed differences by comparing rates of antigenic diffusion to the degree of persistence on phylogeny branches from A/H3N2, A/H1N1, B/Vic and B/Yam (Figure~\ref{accumulation}A).
Here, we observe that in all four clades new antigenic mutations, do not, on average, move antigenic dimension 1 forward or back, consistent with a model of antigenic mutations proceding in a random direction.
However, successful mutations that have persisted in the influenza population, tend to move the antigenic phenotype forward along dimension 1.
This result indicates that forward movement along antigenic dimension 1 is an advantageous trait and is promoted by natural selection \cite{Bedford12}.
The rate of diffusion of persistent lineages is substantially greater in A/H3N2 than in A/H1N1, B/Vic or B/Yam.
For A/H3N2 we see a two-year window during which mutations may be present on side branches of the phylogeny.
After two years, mutations lie along the trunk of the phylogeny and the rate at which the trunk lineage drifts along antigenic dimension 1 is equal to the overall rate at which the population drift along dimension 1.

@@

If we examine the introduction of new mutations in two dimensions, we see that A/H3N2 appears to mutate in antigenic phenotype more rapidly than B/Vic, which mutates more rapidly still than A/H1N1 or B/Yam (Figure~\ref{accumulation}B).
Thus, we have direct evidence that the rate of antigenic mutation is greater in A/H3N2 than in A/H1N1 or influenza B.
However, we observe an interesting additional effect.
The rate of diffusion of persistent lineages is greater than the rate of mutation in influenza A/H3N2, but smaller than the rate of mutation in A/H1N1, B/Vic and B/Yam.
This yields the surprising conclusion that although antigenic mutation is favored in A/H3N2, it is deleterious and opposed by natural selection in A/H1N1 and B.

We suggest that this effect could be due to the pleiotropic effects of antigenic mutations, in which antigenic mutations often result in other phenotypic effects, including decreased protein function \cite{Kaverin04,Rudneva12}.
Thus, we expect that many antigenic variants may be at inherent disadvantage against other strains, even though they benefit from a transmission advantage mediated by host immunity.
Because of this, some antigenic mutations, i.e.\ those that move forward along antigenic dimension 1, may be, on the whole, advantagous, while other antigenic mutations may be, on the whole, deleterious.
In this context, it is especially interesting that antigenic mutations appear to be, on the whole, more advantagous in A/H3N2.
It seems possibe that the H3 protein may have fewer pleiotropic constraints that the the H1 protein or B hemagglutinin protein.
However, more work is necessary to investigate this hypothesis.

@@

Antigenic characteristics of viral strains are often assessed through immunological assays such as hemagglutination inhibition (HI) \cite{Hirst43}.  
At heart, these assays compare the reactivity $H_{ij}$ of one virus strain $i$ to antibodies raised against another virus strain $j$ via challenge or vaccination.  
In the case of HI, the measurement of cross-reactivity takes the form of a titer representing the dilution factor at which serum raised against virus $j$ ceases to be effective at inhibiting binding of virus $i$.  
These factors are commonly assessed by serial dilution, so that HI titers will form a log series, 40, 80, 160, etc \dots.
However, due to experimental constraints, most comparisons cannot be made, leading to a sparse observation matrix $\mathbf{H} = \{H_{ij}\}$.  
Further, measurements are usually interval and truncated, e.g.\ inhibition or neutralization may cease somewhere between the serial titers of 160 and 320, or inhibition or neutralization may be absent at all titers assayed, suggesting a threshold somewhere between 0 and 40.

@@

Let $X_i \in \Re^{P}$ represent the cartographic location of virus $i$ for $i = 1,\ldots,\vn$, and $Y_j$ represent the cartographic location of serum $j$ for $j = 1,\ldots,\sn$.
Virus and antiserum may be isolated from / raised against the same strain and have different cartographic locations.
Typically, $P = 2$, but higher or lower dimensions may better reflect the data.  
This gives set of distances between virus and serum cartographic locations 
\begin{equation}
	\delta_{ij} =  || X_i - Y_j ||_2.
\end{equation}
Here, $H_{ij}$ represents the log$_2$ titer of virus $i$ against serum $j$, and immunological distance can be defined as
\begin{equation}
	d_{ij} =  S_j - H_{ij},
\end{equation}
where $S_j = \max ( H_{1j},\ldots,H_{\vn j} )$.
Traditionally, the goal of multidimensional scaling (MDS) optimizes over $\mathbf{X}$ and $\mathbf{Y}$ such that
\begin{equation}
	\sum_{(i,j) \in \cal I} 
	\left(
		\delta_{ij} - d_{ij}
	\right)^2
\end{equation}
is minimized, where $\cal I = \{ (i,j) : H_{ij} \mbox{ is measured} \}$. 
Here, we instead assume a probabilistic interpretation in which an observed titer is normally distributed around its cartographic expectation
\begin{equation} 
	H_{ij} \sim \mbox{Normal}( S_j - \delta_{ij}, \mdssd^2 ),
\end{equation}
and the likelihood of observing a particular titer given the placement of antigenic locations is 
\begin{equation} 
	\point(H_{ij}) = \phi \left( \frac{ H_{ij} + \delta_{ij} - S_j }{ \mdssd } \right),
\end{equation}
where $\mdssd^2$ represents variance and $\phi(\cdot)$ represents the standard normal PDF.
HI assays sometimes show no inhibition at all measured titrations, e.g.\ a measurement can be reported as `$<$40'.
In this case, the likelihood of observing the threshold measurement follows the cumulative density of the lower tail of the normal distribution
\begin{equation} 
	\threshold(H_{ij}) = \Phi \left( \frac{ H_{ij} + \delta_{ij} - S_j }{ \mdssd } \right),
\end{equation}
where $\Phi(\cdot)$ represents the standard normal CDF.
Although it is simplest to assume that immunological measurements represent point estimates, it seems more natural to assume that the threshold for inhibition occurs between two titers, e.g.\ we observe inhibition at 1:40 dilution and no inhibition at 1:80 dilution.
Rather than taking the HI titer as 40, we can instead treat this as an interval measurement, assuming that the exact titer for inhibition would occur somewhere between 40 and 80.
HI titers are usually reported as the highest titer that successfully inhibits virus binding or growth, so that in this case, we calculate the likelihood of an interval measurement as
\begin{equation} 
	\interval(H_{ij}) = \Phi \left( \frac{ H_{ij} + \delta_{ij} - S_j + 1 }{ \mdssd } \right) - \Phi \left( \frac{ H_{ij} + \delta_{ij} - S_j }{\mdssd} \right).
\end{equation}
These likelihoods are illustrated in Figure~\ref{hij_likelihood}.
Throughout our analyses, we use interval likelihoods $\interval$ rather than point likelihoods $\point$ unless otherwise noted.

@@

We calculate the overall likelihood by multiplying probabilities of individual measurements
\begin{equation} 
	L(\mathbf{X},\mathbf{Y}) = \prod_{(i,j) \in \cal I} f(H_{ij}),
\end{equation}
using probability functions $\point$, $\threshold$ and $\interval$ as appropriate.
We assume that the prior precision $1/\mdssd^2$ follows a diffuse $\mbox{Gamma}(a, b)$ distribution with $a=0.001$ and $b=1000$, and we start by assuming a diffuse normal prior $\mathrm{Normal}(\mu,\Sigma)$ over virus and sera locations $\mathbf{X}$ and $\mathbf{Y}$, where $\mu$ is equal to $\{0,0\}$ and $\Sigma$ is equal to 1000 along diagonal entries and equal to zero along off-diagonal entries.

@@

The preceding model represents immunological distance as a drop in titer against the most reactive comparison for a particular antiserum.  
Here, we relax this assumption and treat the maximum reactivity of a sera as a random variable.
In this case, $H_{ij}$ still follows
\begin{equation} 
	H_{ij} \sim \mbox{Normal}( S_j - \delta_{ij}, \mdssd^2 ),
\end{equation}
but the vector of `serum effects' $\mathbf{S}$ is estimated rather than fixed.
We assume that $S_j$ values are hierarchically distributed according to a normal distribution.
The mean and variance of this distribution is taken from the empirical mean and empirical variance of the set of maximum titers across sera $\{ \max ( H_{1j},\ldots,H_{\vn j} ) : j = 1,\ldots,\sn \}$.
This formulation assumes that particular sera are more reactive in general than other sera.

@@

@@

@@

\pagebreak

\setcounter{figure}{0}
\setcounter{table}{0}
\setcounter{page}{1}
\renewcommand{\thefigure}{S\arabic{figure}}
\renewcommand{\thetable}{S\arabic{table}}
\renewcommand{\thepage}{S\arabic{page}}

%%% SUPPORTING INFORMATION %%%
\section*{Supporting Information}

\pagebreak

>>>>

Before 2009, there were four major clades of influenza circulating within the human population: A/H3N2, A/H1N1, B/Victoria and B/Yamagata. 
In the case of influenza A, subtypes A/H3N2 and A/H1N1 refer to the genes, hemagglutinin (H or HA) and neuraminidase (N or NA), that are primarily responsible for the antigenic character of a strain. 
In the case of influenza B, B/Vic and B/Yam refer to antigenically distinct lineages which diverged from one prior to 1980 \cite{Rota92}.
Mutations to the HA1 region of the hemagglutinin protein are thought to drive the majority of antigenic drift in the influenza virus \cite{Nelson07NatRevGenet}. 
Experimental characterization of antigenic phenotype is possible through the hemagglutination inhibition (HI) assay \cite{Hirst43}, which measures the cross-reactivity of one virus strain to serum raised against another strain through challenge or vaccination. 
Sera from older strains react poorly with more evolved viruses resulting in new strains having a transmission advantage over previously established strains.
HI assays have shown there to be little to no cross-reactivity between influenza clades A/H3N2, A/H1N1, B/Vic and B/Yam \cite{Hay01}.

The results of many HI assays across a multitude of virus strains can be combined to yield a two-dimensional map, quantifying antigenic similarity and distance \cite{Smith04}. 
The antigenic map of influenza A/H3N2 has shown largely linear movement of the influenza virus population since its introduction in 1968. 
Evolution of antigenic phenotype appears punctuated with periods of stasis interspersed by periods of more rapid innovation, while genetic evolution appears more continuous \cite{Smith04}, suggesting that a relatively small number of genetic changes or combinations of genetic changes may drive changes in antigenic phenotype. 
The process of antigenic drift results in the rapid turnover of the virus population. 
Although mutation occurs rapidly, standing genetic diversity is low and phylogenetic analysis shows a characteristically `spindly' tree with a single predominant trunk lineage and transitory side branches that persist for only 1--5 years \cite{Fitch97}.

@@

In order to assess patterns of antigenic evolution among influenza strains, we implemented a Bayesian probabilistic analog of multidimensional scaling (MDS), referred to here as BMDS (see Methods).
In this model, viruses and sera are given $N$-dimensional locations, thus specifying an `antigenic map', such that distances between viruses and sera in this space are inversely proportional to cross-reactivity.
In the BMDS model, a map distance of one antigenic unit translates to an expectation of a 2-fold drop in HI titer between virus and sera.
Maps that produce pairwise distances most congruent with the observed titers will have a high likelihood and will be favored by the BMDS model.
We integrate over sources of uncertainty, such as antigenic locations, in a flexible Bayesian fashion.
We apply this model to HI measurements of virus isolates against post-infection ferret sera for influenza A/H3N2, A/H1N1, B/Vic and B/Yam (see Methods).

We test model performance by constructing a training dataset representing 90\% of the HI measurements of the A/H3N2 dataset and a test dataset representing the remaining 10\% of the full dataset. 
By fitting the BMDS model to the training dataset, we are able to predict HI titers in the test dataset and compare these predicted titers to observed titers in the test dataset. 
We find that a BMDS analog of the model used by Smith et al.\ \cite{Smith04}, in which viruses and sera are represented as 2D locations and expected titer is relative to the maximum titer of a particular ferret serum, performs well, yielding an average absolute predictive error of 0.87 log$_2$ HI titers (Table \ref{errortable}).
We find that a model with two dimensional antigenic locations predicts test data to the same precision as a 3D model, but performs substantially better than a 1D model or models with four or more dimensions (Table \ref{errortable}), and thus specify a two dimensional model in all subsequent analyses.
We extend this model by estimating the strength of overall reactivity of each serum rather than fixing this at the maximum titer, and additionally, by estimating the strength of reactivity of each virus isolate.
We refer to these estimates as serum effects and virus effects, respectively.
We find that including these effects decreases test error to 0.75 for influenza A/H3N2.
We arrive at similar precisions when repeating this analysis for A/H1N1, B/Vic and B/Yam.
For the 2D model with serum and virus effects, A/H1N1 shows a test error of 0.64, B/Vic shows a test error of 0.74 and B/Yam shows a test error of 0.80.

@@

Previous work on influenza antigenic and genetic evolution has examined antigenic and genetic patterns in isolation, and then compared the results to assess congruence of these two aspects of evolution \cite{Hay01,Smith04,Russell08}. 
Here, we simultaneously model antigenic and genetic evolution by adopting an evolutionary diffusion process \cite{Lemey10}, wherein a virus's antigenic character state evolves along branches of phylogenetic tree according to Brownian motion process (see Methods).
The diffusion process acts as a prior, so that genetically similar viruses are expected to have antigenic locations that lie close to one another, and thus we model the relationship between genetic change and accrual of antigenic distance.
We include sequence data for A/H3N2, A/H1N1, B/Vic and B/Yam to estimate this diffusion process.
We find similar levels of predictive accuracy when including the diffusion process with A/H3N2 showing a test error of 0.75, A/H1N1 showing a test error of 0.66, B/Vic showing a test error of 0.70 and B/Yam showing a test error of 0.78.

@@

Through our analysis, we reveal the evolutionary basis for antigenic phenotype in influenza A/H3N2, A/H1N1, B/Vic and B/Yam (Figure~\ref{map}).
Over the time period of 1968 to 2011, influenza A/H3N2 shows substantially more antigenic evolution than is exhibited by A/H1N1 over the course of 1977 to 2009 or B/Vic and B/Yam over the course of 1986 to 2011.
We observe prominent antigenic clusters in A/H3N2 and A/H1N1, but less obvious clustering in B/Vic and B/Yam.
Antigenic clusters show high genetic similarity, so that we only observe a single diffusion event leading to each cluster, rather than the repeated emergence of clusters.
This analysis makes the fate of antigenic clusters obvious, with two clusters in A/H3N2 (Texas/77 and Beijing/89) appearing to be evolutionary dead-ends.

@@

HI assays lack sensitivity beyond a certain point, so that for A/H3N2, cross-reactive measurements only exist between strains sampled at most 11 years apart, leaving only threshold titers, e.g.\ `$<$40', in more temporally distant comparisons.  
Because of the threshold of sensitivity of the HI assay, it's impossible to distinguish a linear trajectory in 2D antigenic space, from a somewhat curved trajectory, so long as the curve does not bring antigenic phenotype full circle to have cross-reactive measurements between temporally distant strains.
To solve this problem of identifiability, we assumed a weak prior that favors linear movement in the 2D antigenic space, with the slope of the linear relationship and the precision of the relationship incorporated into the Bayesian model (see Methods).
Incorporating this prior made no difference in predictive ability of the model; test error for A/H3N2 was 0.75 log$_2$ HI titers whether the prior was included or excluded.
Because of this, we interpret antigenic locations locally rather than globally.
We can determine the rate of antigenic movement of virus lineages without knowing the larger configuration that the movement occurs under.

@@

We find that other clades of influenza evolved in antigenic phenotype substantially slower than A/H3N2 (Figure~\ref{drift}).
Influenza A/H1N1 evolved at a rate of 0.55 units per year (HPD 0.49--0.62), but showing a similar pattern of punctuated antigenic evolution with occasional larger jumps in phenotype, such as the emergence of the Solomon Islands/06 cluster.  
Influenza B/Victoria also evolved relatively slowly, with an average rates of 0.41 (HPD 0.29--0.55) units per year.
Influenza B/Vic appears to show some degree of punctuated antigenic evolution with a recent transition creating the Brisbane/08 cluster.
Interestingly, a minor lineage of B/Vic has persisted, which rather than moving forward in antigenic dimension 1 has moved down in antigenic dimension 2 (Figure~\ref{map}, Figure~\ref{drift}).
The eventual fate of this lineage has yet to be determined.
Influenza B/Yamagata evolved slower still, with an average rate of and 0.21 (HPD 0.10--0.28) units per year.
Very little punctuated evolution is apparent in the evolution of B/Yam.

@@

Additionally, year-to-year comparisons show a similar pattern (Figure~\ref{incidence}). 
Here, we define antigenic drift from year $i-1$ to year $i$ as the difference between the average location in dimension 1 of viruses sampled in year $i-1$ and the average location in dimension 1 of viruses sampled in year $i$.
For instance, we see that 97\% of the antigenic drift between 1998 and 1999 is attributable to A/H3N2 and subsequently the 1999-2000 winter influenza season is dominated by A/H3N2.
Similarly, between 1999 and 2000, most antigenic drift is attributable to A/H1N1 and B/Yam, and the 2000-2001 season is dominated by these same two clades.
Here, we find an overall correlation between clade-specific relative antigenic drift and clade-specific incidence of $r = 0.67$ (Pearson correlation, $p = 1.00 \times 10^{-6}$).
Correlation is present to greater of lesser degrees for each of the four clades, with $r_\mathrm{H3} = 0.63$, $r_\mathrm{H1} = 0.29$, $r_\mathrm{Vic} = 0.75$ and $r_\mathrm{Yam} = 0.58$.

@@

Here, we examine the conversion of newly acquired antigenic mutation into fixed differences by comparing rates of antigenic diffusion to the degree of persistence on phylogeny branches from A/H3N2, A/H1N1, B/Vic and B/Yam (Figure~\ref{accumulation}A).
Here, we observe that in all four clades new antigenic mutations, do not, on average, move antigenic dimension 1 forward or back, consistent with a model of antigenic mutations proceding in a random direction.
However, successful mutations that have persisted in the influenza population, tend to move the antigenic phenotype forward along dimension 1.
This result indicates that forward movement along antigenic dimension 1 is an advantageous trait and is promoted by natural selection \cite{Bedford12}.
The rate of diffusion of persistent lineages is substantially greater in A/H3N2 than in A/H1N1, B/Vic or B/Yam.
For A/H3N2 we see a two-year window during which mutations may be present on side branches of the phylogeny.
After two years, mutations lie along the trunk of the phylogeny and the rate at which the trunk lineage drifts along antigenic dimension 1 is equal to the overall rate at which the population drifts along dimension 1.

@@

If we examine the introduction of new mutations in two dimensions, we see that A/H3N2 appears to mutate in antigenic phenotype more rapidly than B/Vic, which mutates more rapidly still than A/H1N1 or B/Yam (Figure~\ref{accumulation}B).
Thus, we have direct evidence that the rate of antigenic mutation is greater in A/H3N2 than in A/H1N1 or influenza B.
However, we observe an interesting additional effect.
The rate of diffusion of persistent lineages is greater than the rate of mutation in influenza A/H3N2, but smaller than the rate of mutation in A/H1N1, B/Vic and B/Yam.
This yields the surprising conclusion that although antigenic mutation in general is favored in A/H3N2, it is deleterious and opposed by natural selection in A/H1N1 and B.

We suggest that this finding could be due to the pleiotropic effects associated with antigenic change, in which antigenic mutations often result in other phenotypic effects, including decreased protein function \cite{Kaverin04,Rudneva12}.
Thus, we expect that many antigenic variants may be at inherent disadvantage against other strains, even though they benefit from a transmission advantage mediated by host immunity.
Because of this, some antigenic mutations, i.e.\ those that move forward along antigenic dimension 1, may be, on the whole, advantagous, while other antigenic mutations may be, on the whole, deleterious.
In this context, it is especially interesting that antigenic mutations appear to be, on the whole, more advantagous in A/H3N2.
It seems possible that the H3 protein may have fewer pleiotropic constraints that the the H1 protein or B hemagglutinin protein.
However, more work is necessary to investigate this hypothesis.

In this context, it is especially interesting that adaptive fixations in the HA1 protein determined through analysis of nonsynonymous and synonymous evolution is substantially greater in A/H3N2 than in A/H1N1 \cite{Wolf06, Bhatt11}.
Thus, we conclude that additional adaptive fixations at the amino acid level translate to greater rates of adaptive antigenic change.

@@

Antigenic characteristics of viral strains are often assessed through immunological assays such as the hemagglutination inhibition (HI) assay \cite{Hirst43}.  
At heart, these assays compare the reactivity of one virus strain to antibodies raised against another virus strain via challenge or vaccination.  
In the case of HI, the measurement of cross-reactivity takes the form of a titer $H_{ij}$ representing the dilution factor at which serum raised against virus $j$ ceases to be effective at inhibiting binding of virus $i$.  
These factors are commonly assessed by serial dilution, so that HI titers will form a log series, 40, 80, 160, etc \dots.
However, due to experimental constraints, most comparisons cannot be made, leading to a sparse observation matrix $\mathbf{H} = \{H_{ij}\}$.  
Further, measurements are usually interval and truncated, e.g.\ inhibition may cease somewhere between the serial titers of 160 and 320, or inhibition may be absent at all titers assayed, suggesting a threshold somewhere between 0 and 40.

@@

Let $X_i \in \Re^{P}$ represent the cartographic location of virus $i$ for $i = 1,\ldots,\vn$, and $Y_j$ represent the cartographic location of serum $j$ for $j = 1,\ldots,\sn$.
Virus and serum may be isolated from / raised against the same strain and have different cartographic locations.
Typically, $P = 2$, but higher or lower dimensions may better reflect the data.  
This gives a set of distances between virus and serum cartographic locations 
\begin{equation}
	\delta_{ij} =  || X_i - Y_j ||_2.
\end{equation}
With $H_{ij}$ as the log$_2$ titer of virus $i$ against serum $j$, immunological distance can be defined as
\begin{equation}
	d_{ij} =  S_j - H_{ij},
\end{equation}
where $S_j = \max ( H_{1j},\ldots,H_{\vn j} )$.
Traditionally, the goal of multidimensional scaling (MDS) optimizes over $\mathbf{X}$ and $\mathbf{Y}$ such that
\begin{equation}
	\sum_{(i,j) \in \cal I} 
	\left(
		\delta_{ij} - d_{ij}
	\right)^2
\end{equation}
is minimized, where $\cal I = \{ (i,j) : H_{ij} \mbox{ is measured} \}$. 
Here, we instead assume a probabilistic interpretation in which an observed titer is normally distributed around its cartographic expectation
\begin{equation} 
	H_{ij} \sim \mbox{Normal}( S_j - \delta_{ij}, \mdssd^2 ),
\end{equation}
and the likelihood of observing a particular titer given the placement of antigenic locations is 
\begin{equation} 
	\point(H_{ij}) = \phi \left( \frac{ H_{ij} + \delta_{ij} - S_j }{ \mdssd } \right),
\end{equation}
where $\mdssd^2$ represents variance and $\phi(\cdot)$ represents the standard normal PDF.
HI assays sometimes show no inhibition at all measured titrations, e.g.\ a measurement can be reported as `$<$40'.
In this case, the likelihood of observing the threshold measurement follows the cumulative density of the lower tail of the normal distribution
\begin{equation} 
	\threshold(H_{ij}) = \Phi \left( \frac{ H_{ij} + \delta_{ij} - S_j }{ \mdssd } \right),
\end{equation}
where $\Phi(\cdot)$ represents the standard normal CDF.
Although it is simplest to assume that immunological measurements represent point estimates, it seems more natural to assume that the threshold for inhibition occurs between two titers, e.g.\ we observe inhibition at 1:40 dilution and no inhibition at 1:80 dilution.
Rather than taking the HI titer as 40, we can instead treat this as an interval measurement, assuming that the exact titer for inhibition would occur somewhere between 40 and 80.
HI titers are usually reported as the highest titer that successfully inhibits virus binding, so that in this case, we calculate the likelihood of an interval measurement as
\begin{equation} 
	\interval(H_{ij}) = \Phi \left( \frac{ H_{ij} + \delta_{ij} - S_j + 1 }{ \mdssd } \right) - \Phi \left( \frac{ H_{ij} + \delta_{ij} - S_j }{\mdssd} \right).
\end{equation}
These likelihoods are illustrated in Figure~\ref{hij_likelihood}.
Throughout our analyses, we use interval likelihoods $\interval$ rather than point likelihoods $\point$ unless otherwise noted.

@@

We calculate the overall likelihood by multiplying probabilities of individual measurements
\begin{equation} 
	L(\mathbf{X},\mathbf{Y}) = \prod_{(i,j) \in \cal I} f(H_{ij}),
\end{equation}
using probability functions $\point$, $\threshold$ and $\interval$ as appropriate.
We assume that the prior precision $1/\mdssd^2$ follows a diffuse $\mbox{Gamma}(a, b)$ distribution with $a=0.001$ and $b=1000$, and we start by assuming a diffuse normal prior $\mathrm{Normal}(\mu,\Sigma)$ over virus and sera locations $\mathbf{X}$ and $\mathbf{Y}$, where $\mu$ is equal to $\{0,0\}$ and $\Sigma$ is equal to 10,000 along diagonal entries and equal to zero along off-diagonal entries.

@@

The preceding model represents immunological distance as a drop in titer against the most reactive comparison for a particular serum.  
Here, we relax this assumption and treat the maximum reactivity of a sera as a random variable.
In this case, $H_{ij}$ still follows
\begin{equation} 
	H_{ij} \sim \mbox{Normal}( S_j - \delta_{ij}, \mdssd^2 ),
\end{equation}
but the vector of `serum effects' $\mathbf{S}$ is estimated rather than fixed.
We assume that $S_j$ values are hierarchically distributed according to a normal distribution.
The mean and variance of this distribution is taken from the empirical mean and empirical variance of the set of maximum titers across sera $\{ \max ( H_{1j},\ldots,H_{\vn j} ) : j = 1,\ldots,\sn \}$.
This formulation assumes that particular sera are more reactive in general than other sera.

@@

We thank John McCauley for making NIMR vaccine strain selection reports publicly available.
We thank Richard Reeve, Dan Haydon, and Simon Frost for insights on antigenic modeling and MDS.

@@

AR, MAS, TB, and PL conceived the study.
TB, GD, CR, DS and AR gathered antigenic and genetic data.
AR, MAS, TB, and PL designed statistical phylogenetic procedures.
TB performed the analysis.
TB, MAS, AR wrote the paper.

@@

%\pagebreak

%\setcounter{figure}{0}
%\setcounter{table}{0}
%\setcounter{page}{1}
%\renewcommand{\thefigure}{S\arabic{figure}}
%\renewcommand{\thetable}{S\arabic{table}}
%\renewcommand{\thepage}{S\arabic{page}}

%%% SUPPORTING INFORMATION %%%
%\section*{Supporting Information}
